---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load required libraries
```{r load libraries}
library(dplyr)
library(ggplot2)
```

Load data, set up summary data by date and interval to use in the analysis
```{r load and preprocess data}
d <- read.csv("activity/activity.csv", stringsAsFactors = FALSE)
dailySteps <- summarize(group_by(d, date), totalSteps = sum(steps))
intervalSteps <- summarize(group_by(d, interval), avgSteps = mean(steps, na.rm = TRUE))
```

## What is mean total number of steps taken per day?

```{r daily_steps_histogram}
hist(dailySteps$totalSteps, main = "Histogram: Total Steps Per Day", xlab = "Steps")
```

A look at the statistis for the average daily steps data
```{r calculate mean and median for daily steps}
summary(dailySteps)
```


## What is the average daily activity pattern?

Now we will look at the average steps taken in each 5 minute time slice over the course of a day
```{r plot_avg_steps_taken_for_each_interval}
plot(intervalSteps, type = "l", ylab = "Avg. Steps", xlab = "5 Min. Intervals from Midnight", main = "Average Steps per 5 Minute time block per Day")
```

Calculate the time block that has the most average steps
```{r calculate which 5 min interval has the most average steps}
intervalSteps[which.max(intervalSteps$avgSteps),]
```

## Imputing missing values

Calculate the number of NA records in the dataset
```{r calculate and report total number of NA records}
count(d[is.na(d$steps),])
```

For records with missing data, we will use the average value for that interval
```{r create a new dataset with the NA values replaced by the interval mean}
d_noNA <- merge(d, intervalSteps, by.x = "interval", by.y = "interval")
d_noNA <- mutate(d_noNA, steps = ifelse(is.na(steps), avgSteps, steps))
d_noNA <- mutate(d_noNA, avgSteps = NULL)
dailyStepsNoNA <- summarize(group_by(d_noNA, date), totalSteps = sum(steps))
```

Show the average daily steps histogram again, this time with missing values replaced with means
```{r daily_steps_with_NA_replace_histogram}
hist(dailyStepsNoNA$totalSteps, main = "Histogram: Total Steps Per Day with missing values replaced", xlab = "Steps")
```

Recalculate statistics on the average daily steps data with missing values replaced with means
```{r calculate mean and median for daily steps with NA replaced}
summary(dailyStepsNoNA)
```

As you can see, replacing missing values with the time slice average value has little impact on the mean and median.
It does adjust the distribution of values, however, as you can see from the differences in the quantiles.



## Are there differences in activity patterns between weekdays and weekends?


Now we will investigate whether there is a difference in pattern activity on weekends.
First we will create a factor with weekend and weekday levels

```{r add weekday/weekend factor}
d_noNA$date <- as.Date(d_noNA$date)
d_noNA$DayType <- factor((weekdays(d_noNA$date) %in% c('Saturday','Sunday')), levels = c(TRUE,FALSE), labels= c('weekend','weekday'))
intervalStepsNoNA <- summarize(group_by(d_noNA, interval, DayType), avgSteps = mean(steps))
```

```{r plot_weekday_vs_weekend_activity}
ggplot(data = intervalStepsNoNA, aes(x = interval, y = avgSteps)) + geom_line() + facet_wrap(~ DayType, ncol = 1) + ylab("Avg. Steps") + xlab("5 Min Intervals from Midnight")
```

Not surprisingly, there is a noticeably different activity pattern between weekdays and weekends.